The future age structure of scientific disciplines

### 2025-04-16 age_structure_forecasts.qmd
Restructured report to not include physics as special case

# Introduction

The @AAS is dedicated to advancing Australia’s scientific knowledge by providing evidence-informed advice and advocacy to support decision-making in government. As part of this mission, it has launched _Australian Science, Australia’s Future: Science 2035_, an initiative assessing the capability of the national science system and its role in achieving Australia’s ambitions [@AAS2035]. This report supports this initiative by forecasting future science workforce age structures, providing data-driven insights on how Australia’s science workforce will change over the next decade.

This analysis primarily focuses on Natural and Physical Sciences as defined in the Australian Standard Classification of Education (ASCED) by the @ABSASCED (ABS). For consistency, the report refers to ASCED’s Narrow Fields as "disciplines".

@@

The report begins by outlining the [Methodology](#methodology), [Data Sources](#data-sources), and statistical techniques used to generate the forecasts. A [Worked Example](#physics-and-astronomy) using Physics and Astronomy illustrates the application of these methods in detail, providing a foundation for understanding the approach. The [Findings](#findings) section then extends this analysis to all other disciplines, focusing on comparison and contrast rather than procedural details. Finally, the report explores [Broader Applications](#broader-applications) beyond science, considering how these forecasting methods could be applied to other fields. The results aim to inform future workforce planning and policy decisions to support the growth of Australia’s scientific community.

# Methodology {#methodology}

@@

# Data {#data-sources}

@@

Data limitations include the availability of labour force participation data only for 2016 and 2021, the absence of full-time worker data outside census years, and the aggregated nature of retirement intentions data. These gaps are addressed using specific handling of labour force participation data, cohort interpolation methods, and smoothed disaggregation, as detailed in Sections [3.2](#labour), [3.3](#cohort), and [3.4](#smooth).

Another limitation is that retirement intentions data is unavailable by educational discipline, with industry-level data serving as the closest proxy. Since scientists work across multiple industries, this does not capture discipline-specific retirement patterns but is used in this context through methods outlined in Section [3.5](#retirement).

## Labour Force Participation {#labour}

@@

## Cohort Interpolation {#cohort}

@@

## Smoothed Disaggregation {#smooth}

@@

To estimate single-year age distributions, a monotonic cubic spline [@SHW04] is applied directly to the cumulative values of these age groups, ensuring a smooth transition between them. This method preserves the overall distribution while generating continuous single-year estimates, as shown in @fig-afterdisaggregation.

@@

## Retirement Distribution by Age {#retirement}

@@

# Worked Example: Physics and Astronomy {#physics-and-astronomy}

This section provides a detailed demonstration of the forecasting process using Physics and Astronomy as a case study.

The analysis begins by interpolating the working population from available data, followed by the estimation of graduate completions and calculation of net migration. These data points are then used to simulate multiple future scenarios, applying demographic processes of death and retirement. The final step is to compile these components into a projection of the age structure of scientists in Physics and Astronomy for the next 20 years.

This case study establishes the approach, which is later applied across all disciplines for comparison in the [Findings](#findings) section.

@@

To estimate the future working population accurately, historical data is required. The Australian Census provides such data, but it is only available every five years. @fig-physicists shows the population of physicists and astronomers active in the labour market as recorded in the Census for the years 2006, 2011, 2016, and 2021.

Visually, the distribution for each year is positively skewed, with a long right tail. Over time, the distribution has shifted upward, indicating an increase in the total number of working physicists and astronomers. The right tail has gradually grown heavier, reflecting an increasing proportion of older scientists, while the distribution has also become wider, indicating a broader age range in the workforce.

As shown in @tbl-stats-physics, physicists and astronomers have a median age of approximately 44 years, which suggests a concentration of workers in mid-career stages. The 10th and 90th percentiles, representing the typical adult working age range, indicate that the 10th percentile is in the late 20s and the 90th percentile is in the early 60s. This age range has progressively widened from the first to the last Census.

After estimating the intervening years using linear cohort interpolation [@STUPP88], the resulting population estimates are shown in @fig-physicists2. These interpolated values form the basis for forecasting the future working population, representing the values of $P_{i,x,t}$ for physicists and astronomers. The projected changes in age distribution are visible.

```{r}
#| label: fig-physicists
#| fig-cap: The number of scientists in Physics and Astronomy active in the Australian labour market, shown by age, as recorded in the Australian Census for the years 2006, 2011, 2016, and 2021.
#| eval: true
tar_read(fig11)
```

```{r}
#| label: fig-physicists2
#| eval: true
#| fig-cap: Interpolated number of scientists in Physics and Astronomy active in the Australian labour market, 2006–2021, using linear cohort interpolation.
tar_read(fig12)
```

@@

To estimate the number of new graduates entering the workforce, trends in graduate numbers from 2006 to 2023 are examined. As shown in @fig-physics-leavers, the number of graduates majoring in Physics and Astronomy fluctuated around 375 to 500 annually, with no consistent trend, seasonality, or cyclical pattern. These figures represent total graduate completions ($\sum_x N_{i,x,t}$) in Physics and Astronomy.

To forecast future graduate numbers, a global ARIMA model was employed, following the principles outlined by @HM21. ARIMA is a time series forecasting method that identifies underlying trends in past data, assuming these trends will continue. The global model captures overall trends across disciplines by scaling graduate data within each discipline, ensuring proportional contributions from all disciplines before fitting the global ARIMA model. This improves robustness by incorporating information across disciplines. The model is then used to generate discipline-specific forecasts, with simulations run 500 times to account for variability and generate 90% prediction intervals. The results, along with these intervals, are shown in @fig-future-grads-physics for Physics and Astronomy.

```{r}
#| label: fig-physics-leavers
#| eval: true
#| fig-cap: Annual number of graduates in Physics and Astronomy, 2006–2023.
tar_read(fig13)
```

```{r}
#| label: fig-future-grads-physics
#| fig-cap: Projected number of Physics and Astronomy graduates, 2024–2041, with 90% prediction intervals, based on historical data from 2006–2023.
#| eval: true
tar_read(fig14)
```

Since workforce projections require not only total graduate numbers but also their age distribution, the forecasted totals (depicted in @fig-future-grads-physics) are disaggregated using the percentage distribution of graduate completions by age across all disciplines (as shown in @fig-completions2). This enables an estimation of how many graduates will enter each age group in future years, ensuring consistency with observed patterns in the broader graduate population.

@@

Net migration ($G_{i,x,t}$) is estimated using the demographic growth-balance equation, which is rearranged to calculate the net number of migrants each year (@eq-netmigration). @fig-net-migrants shows the age distribution of net migration among scientists in Physics and Astronomy between 2006 and 2020.

@@

```{r}
#| label: fig-net-migrants
#| eval: true
#| fig-cap: Net migration of scientists in Physics and Astronomy by age, 2006–2020.
tar_read(fig15)
```

@@

1.  **Death Probabilities:** Fit a functional data model [@HU07] to the death probabilities shown in @fig-death-probs, with ARIMA models for the coefficients. This model is used to simulate values of age-specific death probabilities in future years.
2.  **Net Migration:** Fit a separate functional data model to the age-specific migrant numbers shown in @fig-net-migrants, with ARIMA models for the coefficients. This model simulates values of age-specific net migrant numbers in future years.
3.  **Graduate Numbers:** Fit an ARIMA model [@fpp3] to simulate total numbers of graduates in each future year from the numbers shown in @fig-physics-leavers. These graduate numbers will then be disaggregated by age using the age distribution shown in @fig-completions2 to estimate age-specific graduate numbers in each year.
4.  **Retirements:** Estimate retirements in each year by multiplying the number of people of working age in each age bracket (@fig-physicists2) by the retirement probabilities shown in @fig-retirements2. Fixed age-specific retirement probabilities are used.
5.  **Deaths:** Estimate deaths in each year by multiplying the number of people of working age in each age bracket by the forecast death probabilities.
6.  **Demographic Growth-Balance:** Use the demographic growth-balance equation [@eq-dgb] to obtain the next year's age-specific population values. This step incorporates the graduate numbers, net migration, retirements, and deaths, to produce the forecasted population for the next year.

@@

The average of the 500 simulations provides the mean age-specific forecast, while quantiles estimate forecast uncertainty. @fig-future-physics-2035 presents the mean and 90% prediction intervals for 2025 and 2035.

@@

Summing over age groups allows for estimating the total number of working scientists in each future year, as shown in @fig-total-future-physics. This highlights whether the workforce is growing, stabilising, or declining. On average, projections indicate continued growth, but at a gradually slower pace. The lower bound remains nearly flat, suggesting workforce stagnation in a conservative scenario. Even in an optimistic scenario, growth only slightly exceeds the current pace.

```{r}
#| label: fig-future-physics-2035
#| eval: true
#| fig-cap: Forecasted age distribution of the working population for Physics and Astronomy in 2025 and 2035, comparing changes over 10 years. Shaded areas represent 90% prediction intervals.
tar_read(fig17)
```

```{r}
#| label: fig-total-future-physics
#| eval: true
#| fig-cap: Forecasted total number of working scientists for Physics and Astronomy (2022–2041), indicating the rate of change. Shaded areas represent 90% prediction intervals.
tar_read(fig18)

```

# Findings: Natural and Physical Sciences {#findings}

The forecasting process applied to Physics and Astronomy is now extended to all scientific disciplines. Rather than reiterating the methodological details, this section presents key findings, highlighting differences and comparisons across disciplines. The analysis follows the same structure, covering the working population, graduate completions, net migration, and forecast results based on future population simulations.

@tbl-disciplines lists the scientific disciplines (ASCED Narrow Fields) included in this analysis, along with their corresponding Detailed Fields. This table provides essential context for understanding the specific areas classified under each discipline. "n.f.d." stands for "Not Further Defined," and "n.e.c." stands for "Not Elsewhere Classified."

```{r}
#| label: tbl-disciplines
#| eval: true
#| tbl-cap: Classification of scientific disciplines, based on the ASCED Narrow Fields of Education within the Broad Field of Natural and Physical Sciences. The table lists their corresponding Detailed Fields.
tar_read(discipline_table)
```

The age distribution of the working population across different scientific disciplines is shown in @fig-working-population. Similar to Physics and Astronomy, the age distributions in these disciplines are positively skewed, with a growing proportion of older workers and a widening age range. Moreover, the number of working scientists has increased over time, especially in Mathematical Sciences in the latest 2021 Census, as well as in Biological Sciences and Other Natural and Physical Sciences. While the Y-axis scales vary across disciplines to reflect differences in population sizes, the data nonetheless indicate a substantial shift in working populations over time within these disciplines.

In terms of age structure, most disciplines have median ages in the mid-forties. However, a few disciplines show lower median ages, including Biological Sciences with an approximate median age of 40, Other Natural and Physical Sciences at 34, and Natural and Physical Sciences (n.f.d.) at 39. A table summarising the statistics for each discipline is provided in @tbl-disciplines-stats in the appendix for further reference.

Additionally, the graph with interpolated data, shown in @fig-interpolated-working-population, provides a smoother representation of trends and highlights changes within each discipline.

```{r}
#| label: fig-interpolated-working-population
#| eval: true
#| fig-cap: "Age distribution of the working population across scientific disciplines, interpolated from Census data between 2006 and 2021. Each panel represents a different discipline, with lines indicating the number of active scientists in the Australian labour market by age and year."
#| fig-height: 6
tar_read(fig20)
```

@fig-graduates-disciplines shows the number of graduates across disciplines from 2006 to 2023. A significant positive trend is observed in three disciplines: Mathematical Sciences, Other Natural and Physical Sciences, and Natural and Physical Sciences (n.f.d.). Note that the Y-axis scales vary across disciplines to reflect differences in the number of completions.

```{r}
#| label: fig-graduates-disciplines
#| eval: true
#| fig-cap: "Number of graduates across scientific disciplines from 2006 to 2023. Each panel represents a different discipline, showing the trend in graduate numbers over time."
tar_read(fig21)
```

Similarly, disciplines such as Biological Sciences, Other Natural and Physical Sciences, and Natural and Physical Sciences (n.f.d.) were characterised by lower median ages in their working populations. The higher graduate numbers in these fields may contribute to this trend, as younger graduates naturally lower the median age of the workforce.

As with the example in Physics and Astronomy, future graduate numbers have been forecasted for each discipline. The corresponding graphs for these forecasts are included in the appendix (see @fig-future-grads-math, @fig-future-grads-chem, @fig-future-grads-earth, @fig-future-grads-bio, @fig-future-grads-other, and @fig-future-grads-nfd).

@fig-net-migration-disciplines shows the age distribution of net migration across disciplines. Emigration during the early career stage is largely driven by international students graduating and returning home. This negative net migration in the early career stage is a significant factor across all disciplines.

@@

The projections indicate wider prediction intervals for younger age groups due to uncertainty in future graduate numbers, while mid-to-late career estimates primarily reflect the aging of existing cohorts. Across disciplines, the typical working population spans from the mid-to-late 20s (10th percentile) to the mid-50s and early 60s (90th percentile), as shown in @tbl-disciplines-stats in the appendix. Since retirements increase after the late 50s, workforce participation beyond 60 serves as a benchmark for identifying trends in delayed retirement and extended career duration.

@@

```{r}
#| label: fig-forecast-total-working-scientists
#| eval: true
#| fig-cap: "Projected total number of working scientists across scientific disciplines from 2022 to 2041. The shaded region represents the 90% prediction interval, the coloured line indicates the mean estimate, and the black line represents historical data."
tar_read(fig24)
```

@@

# Conclusion

@@

>>>>

# Introduction {#sec-intro}

In planning for the future labour market, it is necessary to forecast the age structure of the workforce, in order to enable data-driven decision-making on policies concerning universities and immigration. We demonstrate a statistical modelling approach to this problem, illustrated using various scientific disciplines in Australia, forecasting future workforce age structures over the next decade. The forecasts described have been used by the Australian Academy of Science as part of _Australian Science, Australia’s Future: Science 2035_, an initiative assessing the capability of the national science system and its role in achieving Australia’s ambitions [@AAS2035].

This analysis focuses on the Natural and Physical Sciences as defined in the Australian Standard Classification of Education (ASCED) by the @ABSASCED. For consistency, the report refers to ASCED’s Narrow Fields as "disciplines".

@@

We describe the methodology in @sec-methodology, data sources in @sec-data, and statistical techniques used to generate the forecasts. The results in @sec-results illustrates the application of the ideas to scientific disciplines, focusing on comparison and contrast rather than procedural details. Finally, we provide some discussion and conclusions in @sec-conclusion. The results aim to inform future workforce planning and policy decisions to support the growth of Australia’s scientific community.

# Methodology {#sec-methodology}

@@

# Data {#sec-data}

@@

Data limitations include the availability of labour force participation data only for 2016 and 2021, the absence of full-time worker data outside census years, and the aggregated nature of retirement intentions data. These gaps are addressed using specific handling of labour force participation data, cohort interpolation methods, and smoothed disaggregation, as detailed in Sections -@sec-labour, -@sec-cohort, and -@sec-smooth.

Another limitation is that retirement intentions data is unavailable by educational discipline, with industry-level data serving as the closest proxy. Since scientists work across multiple industries, this does not capture discipline-specific retirement patterns but is used in this context through methods outlined in @sec-retirement.

## Labour Force Participation {#sec-labour}

@@

## Cohort Interpolation {#sec-cohort}

@@

## Smoothed Disaggregation {#sec-smooth}

@@

To estimate single-year age distributions, a monotonic cubic spline [@SHW04] is applied directly to the cumulative values of these age groups, ensuring a smooth transition between them. This method preserves the overall distribution while generating continuous single-year estimates, as shown in @fig-retirements.

@@

## Retirement Distribution by Age {#sec-retirement}

@@

# Results {#sec-results}

@tbl-disciplines lists the scientific disciplines (as per @ABSASCED Narrow Fields) included in this analysis, along with their corresponding Detailed Fields. This table provides essential context for understanding the specific areas classified under each discipline.

```{r}
#| label: tbl-disciplines
#| eval: true
#| tbl-cap: Classification of scientific disciplines, based on the ASCED Narrow Fields of Education within the Broad Field of Natural and Physical Sciences. The table lists their corresponding Detailed Fields.  "n.e.c." stands for "Not Elsewhere Classified."
tar_read(discipline_table)
```

@@

The Australian Census provide working population data every five years. @fig-working-population shows the age distribution of scientists active in the labour market as recorded in the Census for the years 2006, 2011, 2016, and 2021. There are notable differences over time, and between disciplines. The number of working scientists has increased over time, especially in Mathematical Sciences in the latest 2021 Census, as well as in Biological Sciences and Other Natural and Physical Sciences. The right tail has gradually grown heavier, reflecting an increasing proportion of older scientists, while the distribution has also become wider, indicating a broader age range in the workforce.

The working population for years between census are estimated using linear cohort interpolation [@STUPP88], with the resulting population estimates shown in @fig-interpolated-working-population. These interpolated values form the basis for forecasting the future working population, representing the values of $P_{i,x,t}$.

```{r}
#| label: fig-interpolated-working-population
#| eval: true
#| fig-cap: "Age distribution of the working population across scientific disciplines, interpolated from Census data using linear cohort interpolation. Each panel represents a different discipline, with lines indicating the number of active scientists in the Australian labour market by age and year."
#| fig-height: 6
tar_read(fig20)
```

@@

@fig-grad-forecasts shows the number of total graduate completions from 2006 to 2023 ($\sum_x N_{i,x,t}$). A significant positive trend is observed in three disciplines: Mathematical Sciences, Other Natural and Physical Sciences, and Natural and Physical Sciences (n.f.d.). Note that the Y-axis scales vary across disciplines to reflect differences in the number of completions.

To forecast future graduate numbers, a global ARIMA model was employed, following the principles outlined by @HM21. ARIMA is a time series forecasting method that identifies underlying trends in past data, assuming these trends will continue. The global model captures overall trends across disciplines by scaling graduate data within each discipline, ensuring proportional contributions from all disciplines before fitting the global ARIMA model. This improves robustness by incorporating information across disciplines. The model is then used to generate discipline-specific forecasts, with simulations run 500 times to account for variability and generate 90% prediction intervals. The results, along with these intervals, are shown in @fig-grad-forecasts.

Since workforce projections require not only total graduate numbers but also their age distribution, the forecasted totals (depicted in @fig-grad-forecasts) are disaggregated by age using the percentage distribution of graduate completions by age across all disciplines (as shown in @fig-completions2). This enables an estimation of how many graduates will enter each age group in future years, ensuring consistency with observed patterns in the broader graduate population.

@@

Net migration ($G_{i,x,t}$) is estimated using the demographic growth-balance equation, which is rearranged to calculate the net number of migrants each year (@eq-netmigration). @fig-net-migration-disciplines shows the age distribution of net migration among scientists between 2006 and 2020.  Emigration during the early career stage is largely driven by international students graduating and returning home. This negative net migration in the early career stage is a significant factor across all disciplines.

@@

@@

1.  **Death Probabilities:** Fit a functional data model [@HU07] to the death probabilities shown in @fig-death-probs, with ARIMA models for the coefficients. This model is used to simulate values of age-specific death probabilities in future years.
2.  **Net Migration:** Fit a separate functional data model to the age-specific migrant numbers shown in @fig-net-migration-disciplines, with ARIMA models for the coefficients. This model simulates values of age-specific net migrant numbers in future years.
3.  **Graduate Numbers:** Fit an ARIMA model [@fpp3] to simulate total numbers of graduates in each future year from the numbers shown in @fig-grad-forecasts. These graduate numbers will then be disaggregated by age using the age distribution shown in @fig-completions2 to estimate age-specific graduate numbers in each year.
4.  **Retirements:** Estimate retirements in each year by multiplying the number of people of working age in each age bracket (@fig-interpolated-working-population) by the retirement probabilities shown in @fig-retirements2. Fixed age-specific retirement probabilities are used.
5.  **Deaths:** Estimate deaths in each year by multiplying the number of people of working age in each age bracket by the forecast death probabilities.
6.  **Demographic Growth-Balance:** Use the demographic growth-balance equation [@eq-dgb] to obtain the next year's age-specific population values. This step incorporates the graduate numbers, net migration, retirements, and deaths, to produce the forecasted population for the next year.

@@

The average of the 500 simulations provides the mean age-specific forecast, while quantiles estimate forecast uncertainty. @fig-forecast-working-population presents the mean and 90% prediction intervals for 2025 and 2035.

@@

Summing over age groups allows for estimating the total number of working scientists in each future year, as shown in @fig-forecast-total-working-scientists. This highlights whether the workforce is growing, stabilising, or declining. On average, projections indicate continued growth, but at a gradually slower pace. The lower bound remains nearly flat, suggesting workforce stagnation in a conservative scenario. Even in an optimistic scenario, growth only slightly exceeds the current pace.

@@

The projections indicate wider prediction intervals for younger age groups due to uncertainty in future graduate numbers, while mid-to-late career estimates primarily reflect the aging of existing cohorts. Across disciplines, the typical working population spans from the mid-to-late 20s (10th percentile) to the mid-50s and early 60s (90th percentile). Since retirements increase after the late 50s, workforce participation beyond 60 serves as a benchmark for identifying trends in delayed retirement and extended career duration.

@@

```{r}
#| label: fig-forecast-total-working-scientists
#| eval: true
#| fig-cap: "Forecasted total number of working scientists across scientific disciplines from 2022 to 2041. The shaded region represents the 90% prediction interval, the coloured line indicates the mean estimate, and the black line represents historical data."
tar_read(fig24)
```

@@

# Conclusion {#sec-conclusion}

@@

# Acknowledgements

We thank Alexandra Lucchetti from the Australian Academy of Science for sourcing the data required for this project, and for helpful feedback on earlier versions of this report.
